### 2015-07-19 usecase-combinatorial.tex
combinatoric -> combinatorial
Edits through use cases.

\subsection{Combinatoric Naming}
\label{sec:usecase-combinatoric}
%\subsubsection{Gperftools and MPILeaks}

{\tt Gperftools} is a suite of tools from Google that has gained popularity among
developers for its high-performance thread-safe heap and its lightweight profilers.
Unfortunately, two issues made it difficult to maintain {\tt gperftools} installations
at LC.  First, {\tt gperftools} is a C++ library.  Since C++ does not define a standard
ABI, {\tt gperftools} must be re-built with each compiler and compiler version used by client
applications.  Second, building {\tt gperftools} on bleeding-edge architectures
(such as Blue Gene/Q) requires patches and complicated configure lines that
change with each compiler.  One application team tried to maintain their own
builds of {\tt gperftools} for their preferred compilers, but this
task was complicated due to the second issue.

Spack presented a solution to both problems.  Package administrators can use Spack to
easily maintain a central install of {\tt gperftools} across combinations of
compilers and compiler versions.  Spack's {\tt gperftools} package also serves as a central
institutional knowledge repository for builds, as its package files encode
the patches and configure lines required for each platform and compiler combination.
We are investigating whether we can move some of these settings into compiler and
architecture descriptors to further simplify Spack's build template for cross-platform
installations.
\begin{figure}
\begin{minted}[linenos,
               numbersep=5pt,
			   fontsize=\scriptsize,
               frame=lines,
               framesep=2mm]{python}
patch("patch.gpeftools2.4_xlc", when="@2.4 %xlc")

@@

Figure~\ref{fig:gperftools} illustrates per-compiler and platform build rules with
a simplified version of the install routine for {\tt gperftools}.
%(the full {\tt gperftools} install routine includes other compilers and more options).
The package applies a patch if {\tt gperftools} 2.4 is built with the XL compiler,
and it selects the correct configure line based on the spec's platform and compiler.

LC's {\tt mpileaks} tool, which we have used as a running example in this paper,
has an even larger configuration space than {\tt gperftools}.  It must be built
for different compilers, compiler versions, MPI implementations, and MPI implementation
versions. As with {\tt gperftools}, we have been able to install many different
configurations of {\tt mpileaks} using Spack.  Moreover, Spack's virtual dependency system
allows us to
quickly compose a new {\tt mpileaks} build when a new MPI library is deployed at LC, {\it without}
modifying the {\tt mpileaks} package itself.
%Though we simplified its dependency structure
%here, the {\tt callpath} library can be built with three different stack trace libraries:
%{\tt dyninst}, {\tt libunwind}, or {\tt glibc}.  For some codes we may soon need to build
%different versions of {\tt mpileaks} with different {\tt callpath} configurations.

>>>>

\subsection{Combinatorial Naming}
\label{sec:usecase-combinatorial}
%\subsubsection{Gperftools and MPILeaks}

{\tt Gperftools} is a suite of tools from Google that has gained popularity among
developers for its high-performance thread-safe heap and its lightweight profilers.
Unfortunately, two issues made it difficult to maintain {\tt gperftools} installations
at LC.  First, {\tt gperftools} is a C++ library.  Since C++ does not define a standard
ABI, {\tt gperftools} must be rebuilt with each compiler and compiler version used by client applications.  Second, building {\tt gperftools} on bleeding-edge architectures
(such as Blue Gene/Q) requires patches and complicated configure lines that
change with each compiler.  One application team tried to maintain their own
builds of {\tt gperftools}, but the maintenance burden soon became too great.

\begin{figure}
\begin{minted}[linenos,
               numbersep=5pt,
			   fontsize=\scriptsize,
               frame=lines,
               framesep=2mm]{python}
patch("patch.gpeftools2.4_xlc", when="@2.4 %xlc")

@@

Spack presented a solution to both problems.  Package administrators can use Spack to
easily maintain a central install of {\tt gperftools} across combinations of
compilers and compiler versions.  Spack's {\tt gperftools} package also serves as a central
institutional knowledge repository, as its package files encode
the patches and configure lines required for each platform and compiler combination.
%
Figure~\ref{fig:gperftools} illustrates per-compiler and platform build rules with
a simplified version of the install routine for {\tt gperftools}.
%(the full {\tt gperftools} install routine includes other compilers and more options).
The package applies a patch if {\tt gperftools} 2.4 is built with the XL compiler,
and it selects the correct configure line based on the spec's platform and compiler.

LC's {\tt mpileaks} tool, which has been a running example in this paper,
has an even larger configuration space than {\tt gperftools}.  It must be built
for different compilers, compiler versions, MPI implementations, and MPI
versions. As with {\tt gperftools}, we have been able to install many different
configurations of {\tt mpileaks} using Spack.  Moreover, Spack's virtual dependency 
system allows us to quickly compose a new {\tt mpileaks} build when a new MPI library is 
deployed at LC, {\it without} modifying the {\tt mpileaks} package itself.
%Though we simplified its dependency structure
%here, the {\tt callpath} library can be built with three different stack trace libraries:
%{\tt dyninst}, {\tt libunwind}, or {\tt glibc}.  For some codes we may soon need to build
%different versions of {\tt mpileaks} with different {\tt callpath} configurations.

### 2015-07-19 usecase-policy.tex
Change name of tex file to 'spack-paper-sc15'
Edits through use cases.

Spack makes it easy for package maintainers to create and organize package installations.
But it also needs to be easy for end-users to find and use those packages.
Different end-users may have different expectations about how packages should be built and installed,
and those expectations are typically shaped by years of site policies, personal preferences,
and lingering legacy decisions originallly made on a DEC VAX.
Rather than try to dictate these expectations,
Spack allows both end-users and package maintainers to create custom policies dictating
how packages are built and installed at their site.

@@

While Spack can easily create many installations of a package like {\tt mpileaks}
for different compilers and MPI implementations, end users may find Spack's directory layout
confusing, and they may not be able to find the right libraries.
As discussed in Section~\ref{sec:envmodule},
environment modules are commonly used on supercomputer systems to solve this problem,
and Spack allows the package author to automatically create module and dotkit files for packages that Spack installs.

However, even when modules and dotkits are available,
many users will go directly through the filesystem to access package installs.
Spack installs packages into paths based on concretized specs,
which is ideal for maintaining multiple package installations, but may be difficult for an end-user to navigate.
A version of {\tt mpileaks}, for example, may be installed in a location like
%
\begin{minted}[fontsize=\small]{bash}
spack/opt/chaos_5_x86_64_ib/gcc@4.9.2/mpileaks@1.0-db465029
\end{minted}
%
Spack thus allows the creation of views, which are a symbolic-link based directory layout of packages.
Views provide a human-readable directory layout,
which can be taylored to fit alongside existing directory layouts or used to create new layouts.
For example, the above {\tt mpileaks} package may have a view that creates a link in
{\tt /opt/mpileaks-1.0-openmpi} to the Spack installation of {\tt mpilleaks}.
The same package install may be referenced by multiple links and views,
so the above package could also be linked from a more generic {\tt /opt/mpileaks-openmpi}
link (which is useful for users who don't want to hardcode a specific version).
Views can also be used to create symbolic links to specific executables or libraries in an install,
so a Spack-built {\tt gcc 4.9.2} install may have a view that creates links from
{\tt /bin/gcc49} and {\tt /bin/g++49} to the appropriate {\tt gcc} and {\tt g++} executables.

Views are configured through a {\tt spackconfig} file, which can be set up at the site or user level.
For each package or set of packages, the {\tt spackconfig} file contains rules
describing the links that should point into that package.
The link names can be parameterized.
For example, the above {\tt mpileaks} symbolic link might have been created by a rule like:
%
\begin{minted}[fontsize=\small]{bash}
    /opt/${PACKAGE}-${VERSION}-${MPINAME}
\end{minted}
%
On installation or removal,
links are automatically created, deleted, or updated according to these rules.

Spack's views are a projection from a point in a high-dimension space
(the concretized spec, which fully specifies all parameters) to point in a lower-dimension space
(the link name, which may only specify a few parameters).
Thus several package installations may map to the same link.
For example, the above {\tt mpileaks} link could point to an {\tt mpileaks} compiled with
{\tt gcc} or {\tt icc} -- the compiler parameter is not part of the link.
To keep  package installations  consistent and reproducible,
Spack has a well-defined mechanism for resolving conflicting links.
Spack defines a sort order for packages, which arranges packages from ``most desired'' to ``least desired''.
When a symbolic link could resolve to multiple package installations,
Spack points the link at the most desired package.

By default, Spack treats newer versions of packages compiled with newer compilers
as more desirable than older packages built with older compilers.
It has well-defined, but not necessarily meaningful,
preferences for deciding between MPI implementations and different compilers.
A user can override Spack's default sort order through their {\tt spackconfig} file.
For example, at one site users may typically use the Intel compiler,
but some users also use the system's default {\tt gcc 4.4.7}.
These preferences could be stated by adding:
%
\begin{minted}[fontsize=\small]{bash}
    compiler_order = icc,gcc@4.4.7
\end{minted}
%
to the site's {\tt spackconfig} file, which would cause the ambigious {\tt mpileaks} link to point to an {\tt icc} installation.
Any compiler left out of the compiler\_order setting is treated as less preferred than explicitly stated compilers.
Spack can also be configured to treat specific package versions as more preferred than other versions,
which can be useful if a new version is unstable and untested, and should not be advertised with symbolic links.

By default, Spack stores its package files in a mainline repository that is present when users
first run Spack.  At many sites, packages may build sensitive, proprietary software, or they 
may have patches that are not useful outside of a certain company or organization.  Putting
this type of code back into a public repository does not often make sense, and if it makes the
mainline less stable, it can actually make sharing code between sites more difficult.

To support our own private packages, and to support those of LLNL code teams, Spack allows the creation of site-specific variants of packages.  Via the {\tt spackconfig} file, users can specify additional search directories for finding additional {\tt Package} classes.
%
The additional packages are like the {\tt mpileaks} package shown in Figure~\ref{fig:mpileaks}.  However, the extension packages can extend from not only {\tt Package}, but also any of Spack's built-in packages.   Custom packages can inherit from and replace the Spack's default packages, allowing sites to either tweak or completely replace Spack's build recipes.  To continue the previous example, a site can write a {\tt LocalSpindle} Python class, which inherits from Spack's {\tt Spindle} class.  {\tt LocalSpindle} may simply add additional configure flags to the {\tt Spindle} class, while leaving the dependencies and most of the build instructions from its parent class.  For reproducibility, Spack also tracks the {\tt Package} class that drove a specific build.

>>>>

Spack makes it easy to create and organize package installations,
but it must also be easy for end-users to find and use packages on an HPC
system. Different end-users have different expectations about how packages
should be built and installed, and at LLNL those expectations are shaped
by years of site policies, personal preferences, and lingering legacy
decisions originally made on a DEC VAX.
Spack provides mechanisms that allow users to customize it for the needs
of particular sites and users.

%Rather than try to dictate these expectations,
%Spack allows both end-users and package maintainers to create custom policies dictating
%how packages are built and installed at their site.

@@

While Spack can easily create many installations of a package like {\tt mpileaks}
for different compilers and MPI implementations, end users may find Spack's directory layout confusing, and they may not be able to find libraries intuitively.
Spack installs packages into paths based on concretized specs.
This is ideal for maintaining multiple package installations, 
but it may be difficult for an end-user to navigate.
For example, an {\tt mpileaks} installation prefix might be:
%
\begin{minted}[fontsize=\small]{bash}
spack/opt/linux-x86_64/gcc-4.9.2/mpileaks-1.0-db465029
\end{minted}

As discussed in Section~\ref{sec:envmodule},
environment modules are commonly used on HPC systems to solve this problem,
and Spack allows the package author to automatically create environment modules
for packages that Spack installs.
Even when modules are available, many users still
navigate the filesystem to access packages.

Spack thus allows the creation of views, which are a symbolic-link based directory layout of packages.
Views provide a human-readable directory layout,
which can be adapted to match legacy directory layouts that users rely on.
For example, the above {\tt mpileaks} package may have a view that creates a link in
{\tt /opt/mpileaks-1.0-openmpi} to the Spack installation of {\tt mpileaks}.
The same package install may be referenced by multiple links and views,
so the above package could also be linked from a more generic {\tt /opt/mpileaks-openmpi}
link. This is useful for users who always want to use the latest version.
Views can also be used to create symbolic links to specific executables or libraries in an install,
so a Spack-built {\tt gcc@4.9.2} install may have a view that creates links from
{\tt /bin/gcc49} and {\tt /bin/g++49} to the appropriate {\tt gcc} and {\tt g++} executables.

Views are configured using configuration files, which can be set up at the site or
user level.
For each package or set of packages, configuration file contains rules
describing the links that should point into that package.
The link names can be parameterized.
For example, the above {\tt mpileaks} symbolic link might have been created by a rule like:
%
\begin{minted}[fontsize=\small]{bash}
    /opt/${PACKAGE}-${VERSION}-${MPINAME}
\end{minted}
%
On installation and removal,
links are automatically created, deleted, or updated according to these rules.

Spack's views are a projection from a point in a high-dimensional space
(the concretized spec, which fully specifies all parameters) to point
in a lower-dimensional space
(the link name, which may only contain a few parameters).
Thus, several installations may map to the same link.
For example, the above {\tt mpileaks} link could point to an {\tt mpileaks} compiled with
{\tt gcc} or {\tt icc} -- the compiler parameter is not part of the link.
To keep  package installations consistent and reproducible,
Spack has a well-defined mechanism for resolving conflicting links;
it uses a combination of internal default policies and user- or site-defined
policies to define an {\it order} of preference for different parameters.
By default, Spack prefers newer versions of packages compiled with newer compilers
to older packages built with older compilers. It has well-defined, but not 
necessarily meaningful, order of preference for deciding between MPI
implementations and different compilers.
The default policies can be overridden in configuration files, by either users
or by sites. For example, at one site users may typically use the Intel compiler,
but some users also use the system's default {\tt gcc@4.4.7}.
These preferences could be stated by adding:
%
\begin{minted}[fontsize=\small]{bash}
    compiler_order = icc,gcc@4.4.7
\end{minted}
%
to the site's {\tt spackconfig} file, which would cause the ambiguous
{\tt mpileaks} link to point to an installation compiled with {\tt icc}.
Any compiler not in the compiler\_order setting is treated as less preferred
than those explicitly provided.
%
In a similar manner, Spack can be configured to give specific package
configurations priority over others.
This can be useful if a new version is unstable and untested.

\subsubsection{External Package Repositories}
By default, Spack stores its package files in a mainline repository
that is present when users
first run Spack.  At many sites, packages may build sensitive,
proprietary software, or they 
may have patches that are not useful outside of a certain company or
organization.  Putting this type of code back into a public repository
does not often make sense, and if it makes the
mainline less stable, it can actually make sharing code between sites more difficult.

To support our own private packages, and to support those of LLNL code teams,
Spack allows the creation of site-specific variants of packages.
Via configuration files, users can specify additional search
directories for finding additional {\tt Package} classes.
%
The additional packages are like the {\tt mpileaks} package shown in Figure~\ref{fig:mpileaks}.  However, the extension packages can extend from not
only {\tt Package}, but also any of Spack's built-in packages.
Custom packages can inherit from and replace Spack's default
packages, so other sites to either tweak or completely replace Spack's
build recipes.  To continue the previous example, a site can write
a {\tt LocalSpindle} Python class, which inherits from Spack's
{\tt Spindle} class.  {\tt LocalSpindle} may simply add additional
configure flags to the {\tt Spindle} class, while leaving the
dependencies and most of the build instructions from its parent
class.  For reproducibility, Spack also tracks the {\tt Package}
class that drove a specific build.

### 2015-07-19 usecase-python.tex
Change name of tex file to 'spack-paper-sc15'
Edits through use cases.

Python is becoming increasingly popular for HPC applications, 
due to its flexibility as a language and its excellent support
for calling into fast, compiled numerical libraries.
Python is an interpreted language, but one can use it
as a friendlier interface to compiled libraries like FFTW, ATLAS, and
other linear algebra libraries.  Many LLNL code teams use Python in this manner.
%
LC supports Python installations for several application teams,
and maintenance of these repositories has grown increasingly complex over
time.  The problems are similar to those that drove us to create
Spack: different teams want different Python libraries with different
configurations.

Existing Python package managers are either language-specific like 
{\tt easy\_install}~\cite{eby:setuptools}, 
or they do not effectively support building from source~\cite{anaconda,conda}, let
alone multi-configuration builds like Spack's.  More glaringly, the default
model for Python extensions is to install them {\it directly} into the Python
interpreter's prefix, but this makes it impossible to install multiple versions.
An alternative is to install Python extensions in their own prefix,
but then users must manually add each package to their {\tt PYTHONPATH}. 
We wanted a way to  easily manage many different versions and {\it also} allow teams
to have a baseline set of Python packages available by default.

To support this mode of operation, we added the concept of {\tt extension} packages
to Spack. Python modules now use the {\tt extends('python')} directive instead of
{\tt depends\_on('python')}.
Each module installs into its own prefix like any other package,
and each module depends on a particular Python installation.
Additionally, we added the capability to {\tt activate} and {\tt deactivate} an
extension within its dependent Python installation.  When this happens, each file
in the module prefix is symbolically linked into the Python installation prefix, as though it were
directly installed within the Python instance. If any file conflict would arise from this operation,
the activate operation fails and notifies the user.
The deactivate operation removes the symbolic links and restores the Python installation to its 
pristine state.

There were additional complications because many Python packages {\it install their own
package manager} if they do not find {\tt easy\_install} in the Python installation.
There are also no fewer than three ways that Python packages add themselves to the
interpreter's default path, and some of these conflict. We modified Spack so that 
extendable packages, like Python, can supply custom code in their package file that is
executed during activation and deactivation to handle any necessary language-specific tasks.
Python uses this feature to merge conflicting files that are generated by the {\tt easy\_install} tool.

The end result is that Python extensions can now be installed automatically
in their own prefixes, but they can be composed with a wide range
of bleeding-edge libraries that other package managers do not handle.  To experiment
with these extensions, users can load environment modules generated by Spack.
If they want a particular version to be available without any special environment 
settings, they can activate it within the Python instance.

Spack now essentially implements a ``meta package-manager'' for each Python
instance, and it can live together with Spack's normal installation model.
This has allowed us to efficiently support our application teams, for whom we can
now rapidly construct custom Python installations.  We have also reduced
the amount of time that LC staff spend installing Python modules.
Because Spack packages can extend the activation and deactivation mechanisms,
we believe the same mechanism could be used with other 
interpreted languages with similar extension models, such as R, Ruby, or Lua.

>>>>

Python is becoming increasingly popular for HPC applications, 
due to its flexibility as a language and its excellent support
for calling into fast, compiled numerical libraries.
Python is an interpreted language, but one can use it
as a friendlier interface to compiled libraries like FFTW, ATLAS, and
other linear algebra libraries.  Many LLNL code teams use Python in this manner.
%
LC supports Python installations for several application teams,
and maintaining these repositories has grown increasingly complex over
time. The problems are similar to those that drove us to create
Spack: different teams want different Python libraries with different
configurations.

Existing Python package managers are either language-specific~\cite{eby:setuptools}, 
or they do not support building from source~\cite{anaconda,conda}. None
handles multi-configuration builds.  More glaringly, Python extensions
are usually installed into the Python interpreter's prefix.
This makes it impossible to install multiple versions.\footnote{{\tt setuptools}
has support for multiple versions via {\tt pkg_resources}, 
but this requires modifications to client code.}
Installing each extension in its own prefix enables combinatorial versioning,
but it requires users to add packages to the {\tt PYTHONPATH} variable at runtime.

Per Spack's design philosophy, we wanted a way to easily manage many different
versions, but {\it also} to provide a baseline set of extensions {\it without}
requiring environment settings.
%
To support this mode of operation, we added the concept of {\tt extension} packages
to Spack. Python modules use the {\tt extends('python')} directive instead of
{\tt depends\_on('python')}.
Each module installs into its own prefix like any other package,
and each module depends on a particular Python installation.
But extensions can be {\tt activated} or {\tt deactivated} an
within the dependent Python installation.  The {\tt activate} operation
symbolically links each file in the extension prefix into the Python
installation prefix, as though it were installed directly. If any file
conflict would arise from this operation, {\tt activate} fails.
Similarly, the {\tt deactivate} operation removes the symbolic links and restores
the Python installation to its  pristine state.

There were additional complications because many Python packages {\it install their own
package manager} if they do not find one in the Python installation.
There are also many ways that Python packages add themselves to the
interpreter's default path, and some of them conflict. We modified
Spack so that extendable packages, like Python, can supply custom code
in the package file that that specializes {\tt activate} and {\tt deactivate}
for the particular package. Python uses this feature to merge conflicting
files during activation.  The end result is that Python extensions can
be installed automatically in their own prefixes, and they can be composed
with a wide range of bleeding-edge libraries that other package managers do
not handle.

%To experiment with these extensions, users can load environment
%modules generated by Spack. If they want a particular version to be available
%without any special environment settings, they can activate it within the Python instance.

Spack essentially implements a ``meta package-manager'' for each Python
instance, which can coexist with Spack's normal installation model.
This has allowed us to efficiently support our application teams, for whom we can
now rapidly construct custom Python installations.  We have also reduced
the amount of time that LC staff spend installing Python modules.
Because Spack packages can extend the activation and deactivation mechanisms,
we believe the same mechanism could be used with other 
interpreted languages with similar extension models, such as R, Ruby, or Lua.
